---
title: "Assignment 4"
author: "Azoacha Forcheh, 20558994"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: xelatex
  html_document:
    mathjax: local
    self_contained: no
geometry: margin=.75in
graphics: yes
header-includes:
  - \usepackage{graphicx}
  - \usepackage{color}
  - \usepackage{enumitem}
  - \newcommand{\benum}{\begin{enumerate}}
  - \newcommand{\eenum}{\end{enumerate}}
  - \newcommand{\bitem}{\begin{itemize}}
  - \newcommand{\eitem}{\end{itemize}}
fontsize: 9pt
classoption: letter
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

\benum

\item Download the `olive` data from the course website. In that file, there is a dataset on the fatty acid content of 572 different Italian olive oils.  In total eight different fatty acids are measured.  All olive oils come from one of nine different olive growing regions in Italy. 

```{r, echo=FALSE, fig.align="center", fig.width=3}
library(png)
library(grid)
img <- readPNG("./img/OliveRegions.png")
grid.raster(img)
```


Once you load this file into your `R` session there will be a data set called \texttt{olive}. 

```{r, echo=FALSE}
load("./R/olive.Rda")
#head(olive)
```

In this question you are going to focus on the fatty acid `oleic`.
    
\benum
\item (2 marks) Separate the data on `oleic` into 9 different groups as defined by the olive growing `Area`, and draw side by side boxplots of all 9 groups.  Colour the boxplots uniquely using
```{r}
library(colorspace)
cols <- rainbow_hcl(9) # Use these colours.
```

Show your code together with your output.

```{r fig.align="center", fig.width=15, fig.height=15}
groups = with(olive, split(sqrt(oleic), Area))
ord = order(sapply(groups, IQR))
boxplot(groups[ord], col=cols[ord], 
        main="Olive Growing Areas in Italy")
```

\item Load the `R` package `PairViz`.  Use the variate `oleic` and the same colours for the olive growing areas as in part (a) throughout the following:

```{r}
library(PairViz)
```

\benum
\item (3 marks) Suppose we wish every pair of boxplots to appear next to one another in the same plot.
\bitem
\item How many such pairwise comparisons exist?

Since we do not care about the order of the boxplots in the comparisons, the number of ways that we can choose \textbf{2} boxplots to compare out of the total \textbf{9} is $C(9,2) = \frac{9!}{7!2!} = \frac{9(8)}{2} = 36$. Hence, \textbf{36} pairwise comparisons exist.

\item Give the code that will construct this display (without any other constraint on the ordering).
\item Show the display which resulted from your code.

```{r fig.align="center", fig.width=27, fig.height=14}
# Make this bigger and shorten names
n = length(groups)
ordEul = eulerian(n)
boxplot(groups[ordEul], col=cols[ordEul],
        ylab=expression(sqrt("Oleic Level")),
        main="Oleic Levels by Olive Growing Area")
```

\eitem        

\item (5 marks) Suppose we wish every pair of boxplots to appear next to one another in the same plot but that the boxplots should be grouped so that all areas appear only once in each group. 
\bitem
\item Maintaining the same colours for the areas as before, give the code that will construct this display (without any other constraint on the ordering).  
\item Show the display which resulted from your code.

```{r fig.align="center", fig.width=20, fig.height=14}
ordHam = hpaths(n, matrix=FALSE)
boxplot(groups[ordHam], col=cols[ordHam],
        ylab=expression(sqrt("Oleic Level")),
        main="Olive Growing Areas in Italy")
```
\eitem

\item (7 marks) Construct $t$ tests for every pair of olive growing areas (recall  `pairwise.t.test` from class).  Use the significance levels from these tests to construct an ordering of the boxplot pairs, one which favours having the most significantly different pairs at the left of the display.
\bitem
\item Show your code.
\item Show the resulting display.

```{r fig.align="center", fig.width=20, fig.height=14}
# running the pairwise t-tests
t_tests = with(olive,
               pairwise.t.test(sqrt(oleic), Area))

# creating a matrix of weights set to the sig. levels of the t-tests
weights = matrix(0, nrow=n, ncol=n)
rownames(weights) = names(groups)
colnames(weights) = names(groups)
weights[2:n, 1:(n-1)] = t_tests$p.value

# getting rid of the NA values and fixing the weights matrix
diag(weights) = 0
for(i in 1:(n-1)) {
  for(j in (i+1):n) {
    weights[i,j] = weights[j,i]
  }
}

highlowEulOrd = eulerian(-weights)
boxplot(groups[highlowEulOrd], col=cols[highlowEulOrd],
        ylab=expression(sqrt("Oleic Level")),
        main="Olive Growing Areas in Italy")
```

\item Does the ordering perfectly arrange the boxplots so that for any pairwise comparison, those to the left are more significant and those to the right are less significant?

No. The majority of the medians of the boxplots pairs get closer in value, i.e more identical, as the diagram moves from left to right, and the pairs become easier to compare as a result because the boxplots become more in line with each other on the vertical scale. So most of the pairwise comparisons are ordered correctly. However, some of them (e.g. the comparison between Sicily and Coast-Sardinia) are less significant than the comparisons to their right, and should be located further to the right of the plot.

\item Explain why the ordering was successful (or unsuccessful) in this way.

This is likely because a Eulerian path cannot have an edges walked through again. Hence, once the Eulerian has traversed an edge and arrived at a node/region, it will choose to take the next shortest path possible from that node and it cannot travel through edges it has already visited to get to the a different path. Hence, when creating the ordering for the boxplots, the next comparison is determined by the next most significant pairwise comparison for the node that is currently being visited, and this may not always which may not be necessarily the next most significant pairwise comparison overall.

\item Show a display showing only the first 8 comparisons.

```{r fig.align="center", fig.width=15, fig.height=15}
mostSig8 = highlowEulOrd[1:9]

boxplot(groups[mostSig8], 
        col=cols[mostSig8],
        ylab=expression(sqrt("Oleic Level")),
        main="Olive Growing Areas in Italy")
```

\eitem

\item (7 marks) Use the significance levels from part (iii) but now order the boxplots so that the **least** significant differences appear earliest in the sequence from left to right.
\bitem
\item Show your code.
\item Show the resulting display.

```{r fig.align="center", fig.width=20, fig.height=14}
lowhighEulOrd = eulerian(weights)
boxplot(groups[lowhighEulOrd], col=cols[lowhighEulOrd],
        ylab=expression(sqrt("Oleic Level")),
        main="Oleic Levels by Olive Growing Area")
```

\item Does the ordering perfectly arrange the boxplots so that for any pairwise comparison, those to the left are less significant and those to the right are more significant? 

No. For the most part, the medians of the boxplots pairs get further away in value, i.e less identical, as the diagram moves from left to right, and the pairs become more to compare as a result because the boxplots become more out of line with each other on the vertical scale. So most of the pairwise comparisons are ordered correctly. However, some of them (e.g. the comparison between South-Apulia and Coast-Sardinia) are more significant than the comparisons to their right, and should be located further to the right of the plot.

\item Explain why the ordering was successful (or unsuccessful) in this way.

This was unsuccessful for the same reasons as mentioned in the previous question: since the algorithm for `eulerian` is greedy and an edge that has already been traversed cannot be visited again, the next comparison is determined by the next least significant pairwise comparison for the node/region that is currently being visited, and this may not always which may not be necessarily the next least significant pairwise comparison overall.

\item Show a display showing only the first 8 comparisons.

```{r fig.align="center", fig.width=15, fig.height=15}
leastSig8 = lowhighEulOrd[1:9]

boxplot(groups[leastSig8], 
        col=cols[leastSig8],
        ylab=expression(sqrt("Oleic Level")),
        main="Olive Growing Areas in Italy")
```
\eitem

\item (2 marks) Is the sequence used in part (iii) the reverse of that in part (iv)? 

No, it is not.

\bitem
\item If it is, must it be the reverse?  

No, it doesn't because there are 36 unique pairwise comparisons that can be made, so the the set of the 8 most significant ones and the set of the 8 least signifcant ones (but in reverse order) will not be the same, i.e. will not fully overlap, as this could only happen if the Eulerian traversed the same edges again, which by definition does not occur.

\item If it is not, is it impossible for it to be the same? 

For this type of comparison, it is possible.

\item Either way, explain your reasoning.   

Note that the plots of all the comparisons from most significant to least is the reverse of the plots from least to most. Hence, the only way that it would be possible for part (iii) to be the reverse of part (iv) is if there were only 8 pairwise comparisons in total to consider, instead of 36.

\eitem

\eenum

\item The olive growing areas are divided into three different regions: North, South, and Sardinia.  In this part of the question, interest lies only in comparisons between each growing area in the south  and each area in Sardinia.  That is, each southern area (4 areas) is to be compared to each Sardinian area  (2 areas) yielding a total of 8 comparisons of interest.
    
\benum
\item (4 marks)  Having loaded `PairViz`, create a graph having all six areas in the South and Sardinia as nodes and with edges between every pair whose comparison is of interest.
\bitem
\item plot this graph
\item show the code used to create the graph and to plot it.

```{r fig.align="center", fig.width=10, fig.height=10}
regional = olive[olive$Region %in% c('South','Sardinia'),]
regions = with(regional, split(sqrt(oleic), list(Area,Region), drop=TRUE))
nodes = names(regions)
SouthRegions = nodes[endsWith(nodes, "South")]
SardiniaRegions = nodes[!endsWith(nodes, "South")]
RegionGraph = new("graphNEL", nodes=nodes, edgemode="undirected")
for (i in 1:4) {
  stregion = SouthRegions[i]
  for (j in 1:2) {
    sdregion = SardiniaRegions[j]
    RegionGraph = addEdge(stregion, sdregion, RegionGraph)
  }
}

savePar
plot(RegionGraph, "circo")
```
\eitem

\item (4 marks)  Using the graph from part (i), construct an Eulerian and use that Eulerian to produce a sequence of boxplots that show the comparisons of interest.
\bitem
\item show the boxplot display
\item show the code used to construct the Eulerian and the display.

```{r fig.align="center", fig.width=15, fig.height=15}
ssOrd = eulerian(RegionGraph)

colEulOrd = rep(0, length(ssOrd))
for (i in 1:length(ssOrd)) {
  colEulOrd[i] =  which(nodes==ssOrd[i])
}

boxplot(regions[ssOrd], col=cols[colEulOrd],
        ylab=expression(sqrt("Oleic Level")),
        main="Olive Growing Areas in the South and Sardinia")
```

\eitem

\eenum
\eenum

\eenum